Visualizing Independence Using Extended Association Plots

### 2003-07-08 trunk/vcd/paper/dsc2003.Rnw
fixes for dsc-papers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@271 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We propose to use a color-shading similar to the
enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
\cite[a graphical tool for the 
visualization of the observed frequencies in a contingency table; 
see][]{vcd:Hartigan+Kleiner:1984}:
this extension 
uses a color coding of the mosaic tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table.
Positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals can be
classified according to specified shading levels: per default,
residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

>>>>

We propose to use a color-shading similar to the
enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
\cite[a graphical tool for the 
visualization of the observed frequencies in a contingency table; 
see][]{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984}:
this extension uses a color coding of the mosaic tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table.
Positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals can be
classified according to specified shading levels: per default,
residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}. A different
approach for visualizing the residuals of log-linear models is
used by \cite{vcd:Theus+Laurer:1999} who propose to use relative
residuals, i.e., to scale all residuals according to the absolute sum
of residuals. This is necessary when shading
cells in a mosaic plot for expected frequencies, but offers no
advantages when the residuals themselves are visualized.
